---
title: "Adaeze Obinelo's PM566 Final"
execute: 
  echo: false
author: Adaeze Obinelo
format: html
editor: visual
code-fold: true
---

# INVESTIGATION

Full report can be accessed with [this link](https://docs.google.com/document/d/1_Ygh_oVOJOk_on2GcsdoyUcpb-Cj7eJj51bMxuyIQ9M/edit). click above "Download the Report" tab for direct download.

## Background:

Today's political, economic, and social climate seems to be, on a good day, a succession of inconveniences, and on the worst of days, a never-ending cascade of unprecedented challenges and horrors. Especially thanks to recent economic inflation, everyday necessities are taking a greater and greater toll on the average consumer's wallet. Via the consumer price index (CPI) (NBC news, 2023), the overall costs of food, gas and shelter increased significantly over the summer. This increase was certainly felt acutely here in Los Angeles, where gas rose as high as \$6/gallon in some areas, and food prices continued to climb. Purchasing my airfare home this year, I felt the sting on my wallet particularly severely, and I found myself wondering "were airplane tickets always this high?". I am not inexperienced in air travel, having flown frequently prior to college for athletics, during undergrad, and now as an adult to visit family. For some reason, when I think back to purchasing my tickets in the past, I cannot recall feeling the same unenthusiastic and reluctant resignation that I felt this year. For this reason, I thought it would be an interesting investigation to answer the following question; "How has airfare changes in the past few decades?". The present report aims to do exactly this, using the United States Bureau of Transportation Statistics (BTS) data on airfare.

```{r, include=FALSE}

library(dplyr)
library(data.table)
library(tidyverse)
library(ggplot2)
library(readxl)

set1 <- read_excel('1993f.xls')
set2 <- read_excel('2003f.xls')
set3 <-read_excel('2013f.xls')
set4 <- read_excel('2023f.xls')

names(set1) <- set1[1,] # the data loaded with "header" included as data, so fixing that
set1 <- set1[-1,]

names(set2) <- set2[1,] 
set2 <- set2[-1,]

names(set3) <- set3[1,] 
set3 <- set3[-1,]

names(set4) <- set4[1,] 
set4 <- set4[-1,]


names(set1)[names(set1) == "Average Fare ($)"] <- "Avg_1993"
colnames(set1)[7]<- "Adjusted_Avg_Fare_Q2_1993"
colnames(set1)[2]<- "Airport"
colnames(set1)[1]<- "PassRank_1993"
colnames(set1)[8]<- "_1993v2022_Passengers"

names(set2)[names(set2) == "Average Fare ($)"] <- "Avg_2003"
colnames(set2)[7]<- "Adjusted_Avg_Fare_Q2_2003"
colnames(set2)[2]<- "Airport"
colnames(set2)[1]<- "Pass_Rank_2003"
colnames(set2)[8]<- "_2003v2022_Passengers"

names(set3)[names(set3) == "Average Fare ($)"] <- "Avg_2013"
colnames(set3)[7]<- "Adjusted_Avg_Fare_Q2_2013"
colnames(set3)[2]<- "Airport"
colnames(set3)[1]<- "Pass_Rank_2013"
colnames(set3)[8]<- "_2013v2022_Passengers"

names(set4)[names(set4) == "Average Fare ($)"] <- "Avg_2023"
colnames(set4)[7]<- "Adjusted_Avg_Fare_Q2_2023"
colnames(set4)[2]<- "Airport"
colnames(set4)[1]<- "Pass_Rank_2023"
colnames(set4)[8]<- "_2023v2022_Passengers"



```

# Analysis: Am I really paying more to fly home for thanksgiving than anyone ever has before?

[Precise research question:]{.underline} **Is flying round trip from LAX to Logan Airport(BOS) more expensive on average than it has been in previous decades, when adjusting for 2023 inflation?**

Extra: What would be the most expensive airport to fly out of in 2023? What would be the least expensive?

## [About Dataset]{.underline}

The following analysis was conducted using the United States Bureau of Transportation Statistic's (BTS) record of Average Domestic Itinerary Fares. The dataset tabulates average flight fares by airport per quarter, and ranks these fares by a variable called "2022 Passenger Rank", an ordinal variable calculated by comparing the value equivalent to 10% of all passengers served in 2022 (a variable also included in each dataset).

### Source:

You can download the datasets [here](https://www.transtats.bts.gov/averagefare/)

### Update Schedule

Datasets are compiled each financial quarter, and appear to be updated bi-yearly (the most recent dataset is from the 2nd quarter of 2023).

### Quick Look/Summary

1993

```{r}
nrow(set1)

```

```{r}
summary(set1)
```

2003

```{r}
nrow(set2)
```

```{r}
summary(set2)
```

2013

```{r}
nrow(set3)
```

```{r}
summary(set2)
```

2023

```{r}
nrow(set4)
```

```{r}
summary(set2)
```

## [Methods]{.underline}

The following analysis was conducted using the United States Bureau of Transportation Statistic's (BTS) record of Average Domestic Itinerary Fares. This dataset is compiled each financial quarter (Q1-Q4), and the most recent financial quarter data available is from Q2 of 2023. For each quarter, the dataset compiles average flight fares by airport. Included for each airport was a variable called '2022 Passengers', which compares the number of domestic passengers each airport saw for that time period to the number of passengers the same airport sees on average in the year 2022. This variable was used to calculate another variable "2022 Passenger Rank", assigns an ordinal rank to each airport by comparing the value equivalent to 10% of all passengers served in 2022.

[Missing Values]{.underline}: The dataset did not have missing values for average fares, however, it did have values of 0 or 1 for the variable "2022 Passenger Rank". A missing value or a value of 0 for "2022 Passenger Rank" indicates that these airports don't provide commercial domestic flights, so entries with these values were removed (since they would not be airports the consumer is choosing from)

```{r, include=FALSE}

# Filter out records with missing or 0 values in '_1993v2022_Passengers'
set1 <- set1[complete.cases(set1$`_1993v2022_Passengers`) & set1$`_1993v2022_Passengers` != 0, ]

# Filter out records with missing or 0 values in '_2003v2022_Passengers'
set2 <- set2[complete.cases(set2$`_2003v2022_Passengers`) & set2$`_2003v2022_Passengers` != 0, ]

# Filter out records with missing or 0 values in '_2013v2022_Passengers'
set3 <- set3[complete.cases(set3$`_2013v2022_Passengers`) & set3$`_2013v2022_Passengers` != 0, ]

# Filter out records with missing or 0 values in '_2023v2022_Passengers'
set4 <- set4[complete.cases(set4$`_2023v2022_Passengers`) & set4$`_2023v2022_Passengers` != 0, ]


trx1 <- merge(set1, set2, by=c('Airport', 'Airport Name', 'City Name', 'State Name'))
dim(trx1)

trx2 <- merge(set3, set4, by=c('Airport', 'Airport Name', 'City Name', 'State Name'))
dim(trx2)

trx <- merge(trx1, trx2, by=c('Airport', 'Airport Name', 'City Name', 'State Name'))

trx$PassRank_1993 = as.numeric(as.character(trx$PassRank_1993))
trx$Adjusted_Avg_Fare_Q2_1993 = as.numeric(as.character(trx$Adjusted_Avg_Fare_Q2_1993))
trx$Pass_Rank_2003 = as.numeric(as.character(trx$Pass_Rank_2003))
trx$Adjusted_Avg_Fare_Q2_2003 = as.numeric(as.character(trx$Adjusted_Avg_Fare_Q2_2003))

trx$Pass_Rank_2013 = as.numeric(as.character(trx$Pass_Rank_2013))
trx$Adjusted_Avg_Fare_Q2_2013 = as.numeric(as.character(trx$Adjusted_Avg_Fare_Q2_2013))
trx$Pass_Rank_2023 = as.numeric(as.character(trx$Pass_Rank_2023))
trx$Adjusted_Avg_Fare_Q2_2023 = as.numeric(as.character(trx$Adjusted_Avg_Fare_Q2_2023))

trx$Avg_1993 = as.numeric(as.character(trx$Avg_1993))
trx$Avg_2003 = as.numeric(as.character(trx$Avg_2003))
trx$Avg_2013 = as.numeric(as.character(trx$Avg_2013))
trx$Avg_2023 = as.numeric(as.character(trx$Avg_2023))

library(purrr)
trx<-modify_if(trx, ~is.numeric(.), ~round(., 3))


trx <- na.omit(trx)

airports <- read_csv('us-airports.csv')

gps <- airports %>% 
  select(local_code, latitude_deg, longitude_deg)

names(gps) <- gps[1,] 
gps <- gps[-1,]
colnames(gps)[1]<- "Airport"
colnames(gps)[2]<- "lat"
colnames(gps)[3]<- "lon"


trx <- merge(trx, gps, by=c('Airport'))

trx <- trx %>% drop_na(`Airport Name`) #drop rows missing aiport name (from og trx)'

trx$lat = as.numeric(as.character(trx$lat))
trx$lon = as.numeric(as.character(trx$lon))

```

[Coordinate data]{.underline}: For the purpose of additional analysis, coordinate data was taken from [The Humanitarian Data Exchange](https://data.humdata.org/dataset/ourairports-usa?) which compiled a dataset of longitude and latitude values for 508 major US airports. This coordinate dataset was merged with the BTS datasets.

[Validation]{.underline}: Cross-validation was done by comparing the inflation adjusted fare variable for each year (Adjusted_Avg_Fare_Q2) provided originally with my dataset with an dollar to dollar conversion between that year and 2023.

Conversions:

1993: 1-\>2.13

2003: 1-\>1.67

2013: 1-\>1.32

```{r, include=FALSE}
check_indx <- trx %>% 
  select(Avg_1993, Avg_2003, Avg_2013, Avg_2023) %>% 
  mutate(check.adj.93 = (Avg_1993) *2.13,
                  check.adj.03 = (Avg_2003) * 1.67,
                  check.adj.13 = (Avg_2013) * 1.32,
            check.adj.23 = (Avg_2023)* 1)

trx_indx <- merge(trx, check_indx, by=c('Avg_1993', 'Avg_2003', 'Avg_2013', 'Avg_2023'))

```

```{r}

trx_indx %>% 
  select(Adjusted_Avg_Fare_Q2_1993,check.adj.93, Adjusted_Avg_Fare_Q2_2003, check.adj.03, Adjusted_Avg_Fare_Q2_2013, check.adj.13, Adjusted_Avg_Fare_Q2_2023, check.adj.23)%>% 
  slice(1:5)
   
```

The results indicate a very close coincidence between the BTS inflation adjusted fare and our calculated conversion variable, indicating that the dataset is valid with regards to fare cost.

[Analysis]{.underline}: Given that the research question centers around LAX and BOS airports, the overall dataset was queried for trends in relation to these two airports. Due to the great geographic distance between these airports, the overall dataset was stratified by location to better understand region-specific baseline trends. Scatterplots were used to visualize trends between continuous variables. Bar plots were used to visualize continuous variables by categorical variables.

## [Results]{.underline}

```{r}
trx %>%
  group_by(`State Name`) %>%
  summarise(
    Avg_1993 = mean(Avg_1993, na.rm = TRUE),
    Avg_2003 = mean(Avg_2003, na.rm = TRUE),
    Avg_2013 = mean(Avg_2013, na.rm = TRUE),
    Avg_2023 = mean(Avg_2023, na.rm = TRUE))

```

### State with Highest Average Airfare in 1993

```{r}
m1 <- trx %>%
  group_by(`State Name`) %>%
  summarise(
    Avg_1993 = mean(Avg_1993, na.rm = TRUE),
    Avg_2003 = mean(Avg_2003, na.rm = TRUE),
    Avg_2013 = mean(Avg_2013, na.rm = TRUE),
    Avg_2023 = mean(Avg_2023, na.rm = TRUE)
  ) 

m1 %>% 
  arrange(desc(Avg_1993)) %>% 
  select(`State Name`, Avg_1993) %>% 
   slice(1)
  
```

### State with Highest Average Airfare in 2003

```{r}
 m2 <- trx %>%
  group_by(`State Name`) %>%
  summarise(
    Avg_1993 = mean(Avg_1993, na.rm = TRUE),
    Avg_2003 = mean(Avg_2003, na.rm = TRUE),
    Avg_2013 = mean(Avg_2013, na.rm = TRUE),
    Avg_2023 = mean(Avg_2023, na.rm = TRUE)
  ) 
 
 m2 %>% arrange(desc(Avg_2003)) %>% 
  select(`State Name`, Avg_2003) %>% 
   slice(1)
```

### State with Highest Average Airfare in 2013

```{r}
m3 <- trx %>%
  group_by(`State Name`) %>%
  summarise(
    Avg_1993 = mean(Avg_1993, na.rm = TRUE),
    Avg_2003 = mean(Avg_2003, na.rm = TRUE),
    Avg_2013 = mean(Avg_2013, na.rm = TRUE),
    Avg_2023 = mean(Avg_2023, na.rm = TRUE)
  ) 

m3 %>% 
    arrange(desc(Avg_2013)) %>% 
  select(`State Name`, Avg_2013) %>% 
   slice(1)
 
```

### State with Highest Average Airfare in 2023

```{r}
 m4 <-trx %>%
  group_by(`State Name`) %>%
  summarise(
    Avg_1993 = mean(Avg_1993, na.rm = TRUE),
    Avg_2003 = mean(Avg_2003, na.rm = TRUE),
    Avg_2013 = mean(Avg_2013, na.rm = TRUE),
    Avg_2023 = mean(Avg_2023, na.rm = TRUE)
  ) 
  
 m4 %>% arrange(desc(Avg_2023)) %>% 
  select(`State Name`, Avg_2023) %>% 
   slice(1)
```

### Bird's eye view: Comparing Airport Fare to Passenger Rank

```{r}
trx %>% 
  filter(Airport== "LAX"| Airport== "BOS") %>% 
  select(Airport, Pass_Rank_2023, Pass_Rank_2013, Pass_Rank_2003, PassRank_1993) 
  
```

Both LAX and BOS consistently rank among the top 10 for passenger rank for all 4 years.

### Passenger Rank vs Inflation Adjusted fare in the West

```{r, include=FALSE}
names(trx)[names(trx) == "State Name"] <- "State"
west <- trx %>%
  filter(State=="CA"|State=="AK"|State=="WA"|State=="ID"|State=="OR"|State=="NV"|State=="UT")
```

```{r}

west %>%
  ggplot()+
  geom_point(mapping = aes(x = Pass_Rank_2003 , y = Adjusted_Avg_Fare_Q2_2003, color = "2003")) +
  geom_smooth(mapping = aes(x = Pass_Rank_2003 , y = Adjusted_Avg_Fare_Q2_2003, color = "2003"), method = "lm")+
  geom_point(mapping = aes(x = Pass_Rank_2013 , y = Adjusted_Avg_Fare_Q2_2013, color = "2013" )) +
  geom_smooth(mapping = aes(x = Pass_Rank_2013 , y = Adjusted_Avg_Fare_Q2_2013, color = "2013"), method = "lm")+
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Figure 1:Trend of Inflation Adjusted Fare vs Passenger Rank of Western US Airports in 2003")+
  geom_point(mapping = aes(x = PassRank_1993 , y = Adjusted_Avg_Fare_Q2_1993, color = "1993")) +
  geom_smooth(mapping = aes(x = PassRank_1993 , y = Adjusted_Avg_Fare_Q2_1993, color = "1993"), method = "lm")+
  geom_point(mapping = aes(x = Pass_Rank_2023 , y = Adjusted_Avg_Fare_Q2_2023, color = "2023")) +
  geom_smooth(mapping = aes(x = Pass_Rank_2023 , y = Adjusted_Avg_Fare_Q2_2023, color = "2023"), method = "lm")+
  xlab(label = "Passenger Rank") +
  ylab(label = "Inflation Adjusted Average Fare")+
  scale_color_manual(name="legend",values = c("blue","yellow","pink","red"))

  


```

As the Passenger Rank decreases, broadly speaking so does the cost of the average inflation adjusted flight in the West. Additionally, compared to previous years, 2023 fares range lower than the fares from the years prior, excluding a few outliers in the lower passenger rank region of the graph.

### Passenger Rank vs Inflation Adjusted fare in the East

```{r, include=FALSE}

east <- trx %>%
  filter(State=="NY"|State=="PA"|State=="MD"|State=="DE"|State=="NJ"|State=="CT"|State=="RI"|State=="VT"|State=="MA"|State=="NH"|State=="ME")
```

```{r}


east %>%
  ggplot()+
  geom_point(mapping = aes(x = Pass_Rank_2003 , y = Adjusted_Avg_Fare_Q2_2003, color = "2003")) +
  geom_smooth(mapping = aes(x = Pass_Rank_2003 , y = Adjusted_Avg_Fare_Q2_2003, color = "2003"), method = "lm")+
  geom_point(mapping = aes(x = Pass_Rank_2013 , y = Adjusted_Avg_Fare_Q2_2013, color = "2013" )) +
  geom_smooth(mapping = aes(x = Pass_Rank_2013 , y = Adjusted_Avg_Fare_Q2_2013, color = "2013"), method = "lm")+
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Figure 2:Trend of Inflation Adjusted Fare vs Passenger Rank of Eastern US Airports in 2003")+
  geom_point(mapping = aes(x = PassRank_1993 , y = Adjusted_Avg_Fare_Q2_1993, color = "1993")) +
  geom_smooth(mapping = aes(x = PassRank_1993 , y = Adjusted_Avg_Fare_Q2_1993, color = "1993"), method = "lm")+
  geom_point(mapping = aes(x = Pass_Rank_2023 , y = Adjusted_Avg_Fare_Q2_2023, color = "2023")) +
  geom_smooth(mapping = aes(x = Pass_Rank_2023 , y = Adjusted_Avg_Fare_Q2_2023, color = "2023"), method = "lm")+
  xlab(label = "Passenger Rank") +
  ylab(label = "Inflation Adjusted Average Fare")+
  scale_color_manual(name="legend",values = c("blue","yellow","pink","red"))
```

It does not look like this trend of increasing airfare with increasing passenger rank is as strong in the east. Again, 2023 fares range lower than in previous decades when comparing inflation adjusted averages. From these figures, its starting to look like I might actually be paying less to fly out of LAX and BOS, two high traffic airports, than I would be in previous years.

### Geographic Un-adjusted Fare Comparison

```{r}
LAX.BOS <- trx %>% 
  filter(Airport %in% c("LAX", "BOS")) %>% 
  select(Airport, Adjusted_Avg_Fare_Q2_1993, Adjusted_Avg_Fare_Q2_2003, Adjusted_Avg_Fare_Q2_2013, Adjusted_Avg_Fare_Q2_2023, Avg_1993, Avg_2003, Avg_2013, Avg_2023, Pass_Rank_2023, Pass_Rank_2013, Pass_Rank_2003, PassRank_1993)

LAX.BOS %>%
  gather(key = "Year", value = "Unadjusted_Avg_Fare", Avg_1993, Avg_2003, Avg_2013, Avg_2023) %>%
  ggplot(aes(x = Airport, y = Unadjusted_Avg_Fare, fill = Year)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Figure 3: Unadjusted Fare For BOS and LAX by Year")+
  xlab("Airport") +
  ylab("Unadjusted Average Fare") +
  facet_grid(. ~ Year, scales = "free_y", space = "free_y") +
  scale_fill_manual(
    values = c("Avg_1993" = "blue", "Avg_2003" = "yellow", "Avg_2013" = "pink", "Avg_2023" = "red"),
    labels = c("Avg_1993" = "Average Fare 1993", "Avg_2003" = "Average Fare 2003", "Avg_2013" = "Average Fare 2013", "Avg_2023" = "Average Fare 2023")
  ) +
  theme_minimal()

```

Looking specifically at Logan and LAX by year, it appears that the average ticket out of Logan stagnated between 1993 and 2003, then uptrended between 2013 and 2023. The average ticket out of LAX uptrended slightly between 1993 and 2003, then increased more significantly between 2013 and 2023. On average, the price of a ticket out of Logan was higher than a ticket out of LAX in 1993 and 2003, but below a ticket our of LAX in 2013 and 2023.

*How much compared to 1993?*

```{r}

LAX.BOS <- LAX.BOS %>% 
  mutate(
    comp_fare = Adjusted_Avg_Fare_Q2_1993 - Adjusted_Avg_Fare_Q2_2023)

LAX.BOS %>% 
  select(Airport, comp_fare)

278.577	+ 369.449	
```

On average, I would be paying \$648 more for a round trip in 1993 when comparing inflation adjusted rates. So it looks like, at least comparing the years that I did, I actually have it better off with what I pay now compared to what I would be paying in 1993.

## [Conclusion]{.underline}

Evidently, by inflation adjusted values, I am paying less for a trip between LAX and BOS than I have in the past. **This begs the question of why it feels as if I am not.** This can probably be answered best by looking at the unadjusted fare comparisons in Figure 3. Looking specifically at unadjusted fares out of Logan and LAX by year which show that overall, the *unadjusted* fare for airline tickets out of LAX and BOS are higher in 2023 than compared to years prior. While inflation adjusted values provide a good gauge of the "true" cost of a purchase in relation to the strength of the dollar, unadjusted amounts are the "ouch" one feels when actually paying for something in the moment. Since the unadjusted values are higher than they have been in the past, this is likely why it anecdotally feels like airfare prices are going up. The above study does have significant limitations, namely that limitations in man-power limit our analysis to 4 timepoints separated by a decade. Due to this, this analysis definitely misses more acute variation in the dataset, and additionally does not account for the remaining financial quarters of the year.The above study does have significant limitations, namely that limitations in man-power limit our analysis to four time points separated by a decade. Due to this, this analysis definitely misses more acute variation in the dataset, and additionally does not account for the remaining financial quarters of the year. However, limited though it may be, it does provide a bit of assurance to those of us purchasing airfare.

## Cost Map

Please see "Cost Map" tab to view extra maps and the airport with the least and most expensive fare!
