---
title: "Adaeze Obinelo's PM566 Final"
execute: 
  echo: false
author: Adaeze Obinelo
format: html
editor: visual
code-fold: true
---

# INVESTIGATION

## Background:

Today's political, economic, and social climate seems to be, on a good day, a succession of inconveniences, and at worst, a never-ending cascade of unprecedented challenges and horrors. Especially thanks to recent economic inflation, everyday necessities are taking a greater and greater toll on the average consumer's wallet. The overall cost of food, gas and shelter experienced an increase over the summer, measured via the consumer price index (CPI) (NBC news, 2023). This increase was certainly felt acutely here in Los Angeles, where gas rose as high as \$6 in some areas, and food prices continued to climb. Purchasing my airfare home this year, I felt the sting on my wallet, and I found myself wondering "were airplane tickets always this high?". I am not inexperienced in air travel, having flown frequently prior to college for athletics, as well as during undergrad and now as an adult to visit family. For some reason, when I think back to purchasing in the past, I cannot conjure up the same feelings of unenthusiastic and reluctant resignation that I felt this year. For this reason, I thought it would be an interesting investigation; "How has airfare changes in the past few decades". The present report aims do answer this very question, using the United States Bureau of Transportation Statistics (BTS) data on airfare to analyze trends in airfare throughout the past few decades.

```{r, include=FALSE}

library(dplyr)
library(data.table)
library(tidyverse)
library(ggplot2)
library(readxl)

set1 <- read_excel('1993f.xls')
set2 <- read_excel('2003f.xls')
set3 <-read_excel('2013f.xls')
set4 <- read_excel('2023f.xls')

names(set1) <- set1[1,] # the data loaded with "header" included as data, so fixing that
set1 <- set1[-1,]

names(set2) <- set2[1,] 
set2 <- set2[-1,]

names(set3) <- set3[1,] 
set3 <- set3[-1,]

names(set4) <- set4[1,] 
set4 <- set4[-1,]


names(set1)[names(set1) == "Average Fare ($)"] <- "Avg_1993"
colnames(set1)[7]<- "Adjusted_Avg_Fare_Q2_1993"
colnames(set1)[2]<- "Airport"
colnames(set1)[1]<- "PassRank_1993"
colnames(set1)[8]<- "_1993v2022_Passengers"

names(set2)[names(set2) == "Average Fare ($)"] <- "Avg_2003"
colnames(set2)[7]<- "Adjusted_Avg_Fare_Q2_2003"
colnames(set2)[2]<- "Airport"
colnames(set2)[1]<- "Pass_Rank_2003"
colnames(set2)[8]<- "_2003v2022_Passengers"

names(set3)[names(set3) == "Average Fare ($)"] <- "Avg_2013"
colnames(set3)[7]<- "Adjusted_Avg_Fare_Q2_2013"
colnames(set3)[2]<- "Airport"
colnames(set3)[1]<- "Pass_Rank_2013"
colnames(set3)[8]<- "_2013v2022_Passengers"

names(set4)[names(set4) == "Average Fare ($)"] <- "Avg_2023"
colnames(set4)[7]<- "Adjusted_Avg_Fare_Q2_2023"
colnames(set4)[2]<- "Airport"
colnames(set4)[1]<- "Pass_Rank_2023"
colnames(set4)[8]<- "_2023v2022_Passengers"



```

# Research Question: Am I really paying more to fly home for thanksgiving than anyone ever has before?

Every year it seems like buying my ticket home for the holidays hurts a little bit more than it did the year before. Am I being dramatic, or is the cost of airfare actually going up compared to years prior?

[Precise research question:]{.underline} **Is flying round trip from LAX to Logan Airport(BOS) more expensive on average than it has been in previous decades, when adjusting for 2023 inflation?**

Extra: What would be the most expensive airport to fly out of in 2023? What would be the least expensive?

## [About Dataset]{.underline}

The following analysis was conducted using the United States Bureau of Transportation Statistic's (BTS) record of Average Domestic Itinerary Fares. The dataset compiles average flight fares by airport by year, and ranks these fares by a variable called "2022 Passenger Rank", which is an ordinal variable calculated by comparing the value equivalent to 10% of all passengers served in 2022 (a variable also included in each dataset).

### Source:

You can download the datasets [here](https://www.transtats.bts.gov/averagefare/)

### Update Schedule

Datasets are compiled each financial quarter, and appear to be updated bi-yearly (the most recent dataset is from the 2nd quarter of 2023).

### Quick Look/Summary

1993

```{r}
nrow(set1)

```

```{r}
summary(set1)
```

2003

```{r}
nrow(set2)
```

```{r}
summary(set2)
```

2013

```{r}
nrow(set3)
```

```{r}
summary(set2)
```

2023

```{r}
nrow(set4)
```

```{r}
summary(set2)
```

## [Methods]{.underline}

Due to the data limitations for the current financial year, data from the second financial quarter of 4 years, 1993, 2003, 2013, and 2023 was utilized for analysis. A 10 year spread was thought to be ideal for assessing the general trends in airfare data. The datasets from this chosen time frames were downloaded directly from the Transportation Bureau's website, using the above link, into .csv filed.

As part of data processing, variables names were altered to aid in the coding process. Variable types were changed to the appropriate type (continuous or character).

[Missing Values]{.underline}: Included in each dataset was '2022 Passengers', a variable which compares the amount of domestic passengers each airport sees in 2022. A missing value or a value of 0 indicate these airports don't provide commercial domestic flights, so entries with these values were removed since they would not be airports the consumer is choosing from.

```{r, include=FALSE}

# Filter out records with missing or 0 values in '_1993v2022_Passengers'
set1 <- set1[complete.cases(set1$`_1993v2022_Passengers`) & set1$`_1993v2022_Passengers` != 0, ]

# Filter out records with missing or 0 values in '_2003v2022_Passengers'
set2 <- set2[complete.cases(set2$`_2003v2022_Passengers`) & set2$`_2003v2022_Passengers` != 0, ]

# Filter out records with missing or 0 values in '_2013v2022_Passengers'
set3 <- set3[complete.cases(set3$`_2013v2022_Passengers`) & set3$`_2013v2022_Passengers` != 0, ]

# Filter out records with missing or 0 values in '_2023v2022_Passengers'
set4 <- set4[complete.cases(set4$`_2023v2022_Passengers`) & set4$`_2023v2022_Passengers` != 0, ]


trx1 <- merge(set1, set2, by=c('Airport', 'Airport Name', 'City Name', 'State Name'))
dim(trx1)

trx2 <- merge(set3, set4, by=c('Airport', 'Airport Name', 'City Name', 'State Name'))
dim(trx2)

trx <- merge(trx1, trx2, by=c('Airport', 'Airport Name', 'City Name', 'State Name'))

trx$PassRank_1993 = as.numeric(as.character(trx$PassRank_1993))
trx$Adjusted_Avg_Fare_Q2_1993 = as.numeric(as.character(trx$Adjusted_Avg_Fare_Q2_1993))
trx$Pass_Rank_2003 = as.numeric(as.character(trx$Pass_Rank_2003))
trx$Adjusted_Avg_Fare_Q2_2003 = as.numeric(as.character(trx$Adjusted_Avg_Fare_Q2_2003))

trx$Pass_Rank_2013 = as.numeric(as.character(trx$Pass_Rank_2013))
trx$Adjusted_Avg_Fare_Q2_2013 = as.numeric(as.character(trx$Adjusted_Avg_Fare_Q2_2013))
trx$Pass_Rank_2023 = as.numeric(as.character(trx$Pass_Rank_2023))
trx$Adjusted_Avg_Fare_Q2_2023 = as.numeric(as.character(trx$Adjusted_Avg_Fare_Q2_2023))

trx$Avg_1993 = as.numeric(as.character(trx$Avg_1993))
trx$Avg_2003 = as.numeric(as.character(trx$Avg_2003))
trx$Avg_2013 = as.numeric(as.character(trx$Avg_2013))
trx$Avg_2023 = as.numeric(as.character(trx$Avg_2023))

library(purrr)
trx<-modify_if(trx, ~is.numeric(.), ~round(., 3))


trx <- na.omit(trx)

airports <- read_csv('us-airports.csv')

gps <- airports %>% 
  select(local_code, latitude_deg, longitude_deg)

names(gps) <- gps[1,] 
gps <- gps[-1,]
colnames(gps)[1]<- "Airport"
colnames(gps)[2]<- "lat"
colnames(gps)[3]<- "lon"


trx <- merge(trx, gps, by=c('Airport'))

trx <- trx %>% drop_na(`Airport Name`) #drop rows missing aiport name (from og trx)'

trx$lat = as.numeric(as.character(trx$lat))
trx$lon = as.numeric(as.character(trx$lon))

```

[Coordinate data]{.underline}: For the purpose of additional analysis, coordinate data was taken from [The Humanitarian Data Exchange](https://data.humdata.org/dataset/ourairports-usa?) which compiled a dataset of longitude and latitude values for 508 major US airports. This coordinate dataset was merged with the BTS datasets.

[Validation]{.underline}: Cross-validation was done by comparing the inflation adjusted fare variable for each year (Adjusted_Avg_Fare_Q2) provided originally with my dataset with an dollar to dollar conversion between that year and 2023.

Conversions:

1993: 1-\>2.13

2003: 1-\>1.67

2013: 1-\>1.32

```{r, include=FALSE}
check_indx <- trx %>% 
  select(Avg_1993, Avg_2003, Avg_2013, Avg_2023) %>% 
  mutate(check.adj.93 = (Avg_1993) *2.13,
                  check.adj.03 = (Avg_2003) * 1.67,
                  check.adj.13 = (Avg_2013) * 1.32,
            check.adj.23 = (Avg_2023)* 1)

trx_indx <- merge(trx, check_indx, by=c('Avg_1993', 'Avg_2003', 'Avg_2013', 'Avg_2023'))

```

```{r}

trx_indx %>% 
  select(Adjusted_Avg_Fare_Q2_1993,check.adj.93, Adjusted_Avg_Fare_Q2_2003, check.adj.03, Adjusted_Avg_Fare_Q2_2013, check.adj.13, Adjusted_Avg_Fare_Q2_2023, check.adj.23)%>% 
  slice(1:5)
   
```

The results indicate a very close coincidence between the BTS inflation adjusted fare and our calculated conversion variable, indicating that the dataset is valid with regards to fare cost.

[Analysis]{.underline}: Given that the research question centers around LAX and BOS airports, the overall dataset was queried for trends in relation to these two airports. Due to the great geographic distance between these airports, the overall dataset was stratified by location to better understand region-specific baseline trends. Scatterplots were used to visualize trends between continuous variables. Bar plots were used to visualize continuous variables by geographic location.

## [Results]{.underline}

```{r}
trx %>%
  group_by(`State Name`) %>%
  summarise(
    Avg_1993 = mean(Avg_1993, na.rm = TRUE),
    Avg_2003 = mean(Avg_2003, na.rm = TRUE),
    Avg_2013 = mean(Avg_2013, na.rm = TRUE),
    Avg_2023 = mean(Avg_2023, na.rm = TRUE))

```

### State with Highest Average Airfare in 1993

```{r}
m1 <- trx %>%
  group_by(`State Name`) %>%
  summarise(
    Avg_1993 = mean(Avg_1993, na.rm = TRUE),
    Avg_2003 = mean(Avg_2003, na.rm = TRUE),
    Avg_2013 = mean(Avg_2013, na.rm = TRUE),
    Avg_2023 = mean(Avg_2023, na.rm = TRUE)
  ) 

m1 %>% 
  arrange(desc(Avg_1993)) %>% 
  select(`State Name`, Avg_1993) %>% 
   slice(1)
  
```

### State with Highest Average Airfare in 2003

```{r}
 m2 <- trx %>%
  group_by(`State Name`) %>%
  summarise(
    Avg_1993 = mean(Avg_1993, na.rm = TRUE),
    Avg_2003 = mean(Avg_2003, na.rm = TRUE),
    Avg_2013 = mean(Avg_2013, na.rm = TRUE),
    Avg_2023 = mean(Avg_2023, na.rm = TRUE)
  ) 
 
 m2 %>% arrange(desc(Avg_2003)) %>% 
  select(`State Name`, Avg_2003) %>% 
   slice(1)
```

### State with Highest Average Airfare in 2013

```{r}
m3 <- trx %>%
  group_by(`State Name`) %>%
  summarise(
    Avg_1993 = mean(Avg_1993, na.rm = TRUE),
    Avg_2003 = mean(Avg_2003, na.rm = TRUE),
    Avg_2013 = mean(Avg_2013, na.rm = TRUE),
    Avg_2023 = mean(Avg_2023, na.rm = TRUE)
  ) 

m3 %>% 
    arrange(desc(Avg_2013)) %>% 
  select(`State Name`, Avg_2013) %>% 
   slice(1)
 
```

### State with Highest Average Airfare in 2023

```{r}
 m4 <-trx %>%
  group_by(`State Name`) %>%
  summarise(
    Avg_1993 = mean(Avg_1993, na.rm = TRUE),
    Avg_2003 = mean(Avg_2003, na.rm = TRUE),
    Avg_2013 = mean(Avg_2013, na.rm = TRUE),
    Avg_2023 = mean(Avg_2023, na.rm = TRUE)
  ) 
  
 m4 %>% arrange(desc(Avg_2023)) %>% 
  select(`State Name`, Avg_2023) %>% 
   slice(1)
```

### Bird's eye view: Comparing Airport Fare to Passenger Rank

```{r}
trx %>% 
  filter(Airport== "LAX"| Airport== "BOS") %>% 
  select(Airport, Pass_Rank_2023, Pass_Rank_2013, Pass_Rank_2003, PassRank_1993) 
  
```

Both LAX and BOS consistently rank among the top 10 for passenger rank for all 4 years.

### Passenger Rank vs Inflation Adjusted fare in the West

```{r, include=FALSE}
names(trx)[names(trx) == "State Name"] <- "State"
west <- trx %>%
  filter(State=="CA"|State=="AK"|State=="WA"|State=="ID"|State=="OR"|State=="NV"|State=="UT")
```

```{r}

west %>%
  ggplot()+
  geom_point(mapping = aes(x = Pass_Rank_2003 , y = Adjusted_Avg_Fare_Q2_2003, color = "2003")) +
  geom_smooth(mapping = aes(x = Pass_Rank_2003 , y = Adjusted_Avg_Fare_Q2_2003, color = "2003"), method = "lm")+
  geom_point(mapping = aes(x = Pass_Rank_2013 , y = Adjusted_Avg_Fare_Q2_2013, color = "2013" )) +
  geom_smooth(mapping = aes(x = Pass_Rank_2013 , y = Adjusted_Avg_Fare_Q2_2013, color = "2013"), method = "lm")+
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Figure 1:Trend of Inflation Adjusted Fare vs Passenger Rank of Western US Airports in 2003")+
  geom_point(mapping = aes(x = PassRank_1993 , y = Adjusted_Avg_Fare_Q2_1993, color = "1993")) +
  geom_smooth(mapping = aes(x = PassRank_1993 , y = Adjusted_Avg_Fare_Q2_1993, color = "1993"), method = "lm")+
  geom_point(mapping = aes(x = Pass_Rank_2023 , y = Adjusted_Avg_Fare_Q2_2023, color = "2023")) +
  geom_smooth(mapping = aes(x = Pass_Rank_2023 , y = Adjusted_Avg_Fare_Q2_2023, color = "2023"), method = "lm")+
  xlab(label = "Passenger Rank") +
  ylab(label = "Inflation Adjusted Average Fare")+
  scale_color_manual(name="legend",values = c("blue","yellow","pink","red"))

  


```

As the Passenger Rank decreases, broadly speaking so does the cost of the average inflation adjusted flight in the West. Additionally, compared to previous years, 2023 fares range lower than the fares from the years prior, excluding a few outliers in the lower passenger rank region of the graph.

### Passenger Rank vs Inflation Adjusted fare in the East

```{r, include=FALSE}

east <- trx %>%
  filter(State=="NY"|State=="PA"|State=="MD"|State=="DE"|State=="NJ"|State=="CT"|State=="RI"|State=="VT"|State=="MA"|State=="NH"|State=="ME")
```

```{r}


east %>%
  ggplot()+
  geom_point(mapping = aes(x = Pass_Rank_2003 , y = Adjusted_Avg_Fare_Q2_2003, color = "2003")) +
  geom_smooth(mapping = aes(x = Pass_Rank_2003 , y = Adjusted_Avg_Fare_Q2_2003, color = "2003"), method = "lm")+
  geom_point(mapping = aes(x = Pass_Rank_2013 , y = Adjusted_Avg_Fare_Q2_2013, color = "2013" )) +
  geom_smooth(mapping = aes(x = Pass_Rank_2013 , y = Adjusted_Avg_Fare_Q2_2013, color = "2013"), method = "lm")+
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Figure 2:Trend of Inflation Adjusted Fare vs Passenger Rank of Eastern US Airports in 2003")+
  geom_point(mapping = aes(x = PassRank_1993 , y = Adjusted_Avg_Fare_Q2_1993, color = "1993")) +
  geom_smooth(mapping = aes(x = PassRank_1993 , y = Adjusted_Avg_Fare_Q2_1993, color = "1993"), method = "lm")+
  geom_point(mapping = aes(x = Pass_Rank_2023 , y = Adjusted_Avg_Fare_Q2_2023, color = "2023")) +
  geom_smooth(mapping = aes(x = Pass_Rank_2023 , y = Adjusted_Avg_Fare_Q2_2023, color = "2023"), method = "lm")+
  xlab(label = "Passenger Rank") +
  ylab(label = "Inflation Adjusted Average Fare")+
  scale_color_manual(name="legend",values = c("blue","yellow","pink","red"))
```

It does not look like this trend of increasing airfare with increasing passenger rank is as strong in the east. Again, 2023 fares range lower than in previous decades when comparing inflation adjusted averages. From these figures, its starting to look like I might actually be paying less to fly out of LAX and BOS, two high traffic airports, than I would be in previous years.

### Geographic Un-adjusted Fare Comparison

```{r}
LAX.BOS <- trx %>% 
  filter(Airport %in% c("LAX", "BOS")) %>% 
  select(Airport, Adjusted_Avg_Fare_Q2_1993, Adjusted_Avg_Fare_Q2_2003, Adjusted_Avg_Fare_Q2_2013, Adjusted_Avg_Fare_Q2_2023, Avg_1993, Avg_2003, Avg_2013, Avg_2023, Pass_Rank_2023, Pass_Rank_2013, Pass_Rank_2003, PassRank_1993)

LAX.BOS %>%
  gather(key = "Year", value = "Unadjusted_Avg_Fare", Avg_1993, Avg_2003, Avg_2013, Avg_2023) %>%
  ggplot(aes(x = Airport, y = Unadjusted_Avg_Fare, fill = Year)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Figure 3: Unadjusted Fare For BOS and LAX by Year")+
  xlab("Airport") +
  ylab("Unadjusted Average Fare") +
  facet_grid(. ~ Year, scales = "free_y", space = "free_y") +
  scale_fill_manual(
    values = c("Avg_1993" = "blue", "Avg_2003" = "yellow", "Avg_2013" = "pink", "Avg_2023" = "red"),
    labels = c("Avg_1993" = "Average Fare 1993", "Avg_2003" = "Average Fare 2003", "Avg_2013" = "Average Fare 2013", "Avg_2023" = "Average Fare 2023")
  ) +
  theme_minimal()

```

Looking specifically at Logan and LAX by year, it appears that the average ticket out of Logan stagnated between 1993 and 2003, then uptrended between 2013 and 2023. The average ticket out of LAX uptrended slightly between 1993 and 2003, then increased more significantly between 2013 and 2023. On average, the price of a ticket out of Logan was higher than a ticket out of LAX in 1993 and 2003, but below a ticket our of LAX in 2013 and 2023.

*How much compared to 1993?*

```{r}

LAX.BOS <- LAX.BOS %>% 
  mutate(
    comp_fare = Adjusted_Avg_Fare_Q2_1993 - Adjusted_Avg_Fare_Q2_2023)

LAX.BOS %>% 
  select(Airport, comp_fare)

278.577	+ 369.449	
```

On average, I would be paying \$648 more for a round trip in 1993 when comparing inflation adjusted rates. So it looks like, at least comparing the years that I did, I actually have it better off with what I pay now compared to what I would be paying in 1993.

## [Conclusion]{.underline}

It looks like if we're just looking at the dollar amount, yes I spend slightly more in 2023 than what people spent for the average round trip from BOS to LAX in previous years, primarily thanks to the rising cost of a ticket out of LAX. However, in inflation adjusted dollars (which is more relevant, even though it doesn't feel any better) I am actually spending less. [So in conclusion, I am **not** spending more in relative 2023 dollars than anyone else has every spent before for a round trip between LAX and Boston.]{.underline} The above study does have significant limitations, namely that limitations in man-power limit our analysis to 4 timepoints separated by a decade. Due to this, this analysis definitely misses more acute variation in the dataset, and additionally does not account for the remaining financial quarters of the year. However, limited though it may be, it does provide a bit of assurance to those of us purchasing airfare.

## Cost Map

Please see "Cost Map" tab to view extra maps and the airport with the least and most expensive fare!
